Periodograms for Multiband Astronomical Time Series

### 2015-02-03 writeup/paper.tex
JTV final pass

\begin{abstract}
This paper introduces the {\it multiband periodogram}, a general extension of the standard 
Lomb-Scargle approach for detecting periodic signals in time-domain data. In addition to 
advantanges of the Lomb-Scargle method such as treatment of non-uniform sampling and
heteroscedastic errors, the multiband periodogram significantly improves period finding for randomly 
sampled multiband light curves (e.g., Pan-STARRS, DES and LSST). The light curves in 
each band are modeled as arbitrary truncated Fourier series, with the period and phase 
shared across all bands. The key aspect is the use of Tikhonov regularization which drives
most of the variability into the so-called base model common to all bands, while 
fits for individual bands describe residuals relative to the base model and typically require
lower-order Fourier series. This decrease in the required number of fit parameters is the
main reason for improved performance. We use simulated light curves 
and randomly subsampled SDSS Stripe 82 data to demonstrate the superiority of this 
method compared to other methods from the literature, and find that this method will be able to
efficiently determine the correct period in the majority of LSST's bright RR Lyrae stars with
as little as six months of LSST data.
The Python implementation of this method, along with code to fully reproduce the results
reported here, is available on GitHub.
\end{abstract}

@@

The proposed method is essentially a generalization of the Lomb-Scargle method to 
multiband case. The light curves in  each band are modeled as arbitrary truncated Fourier series, 
with the period, and optionally the phase, shared across all bands. The key aspect enabling this approach is the use of Tikhonov regularization 
(discussed in detail in \S \ref{regularization}) which drives most of the variability into the so-called {\it base 
model} common to all bands, while fits for individual bands describe residuals relative to the base model 
and typically require lower-order Fourier series. This decrease in the required number of fit parameters is the
main reason for improved performance.

In \sect{brief_overview} we provide a brief overview of least-squares periodic fitting, and in \sect{matrix_formalism} derive the matrix-based formalism for single-band periodic analysis used through the rest of this work.
\sect{extending_periodogram} introduces several extensions and generalizations to the single-band model that the matrix formalism makes possible, including floating mean models, truncated Fourier models, and regularized models.
In \sect{multiband}, we use the ideas behind these extensions to motivate the {\it multiband periodogram}, and show some examples of its use on simulated data.
In \sect{stripe82} we apply this method to measurements of 483 RR Lyrae stars first explored by \citet[][hereafter S10]{Sesar2010}, and conclude in \sect{discussion}.

\section{Brief Overview of Periodograms}
\sectlabel{brief_overview}

The detection and quantification of periodicity in time-varying signals is an important area of data analysis within modern time-domain astronomical surveys.
For evenly-spaced data, the {\it Schuster periodogram}, introduced in 1905, gives a quantitative measure of the periodicity of data as a function of the angular frequency $\omega$. For data $\{y_k\}_{k=1}^N$ measured at equal intervals $t_k = t_0 + k\Delta t$, the Schuster periodogram, which measures the spectral power as a function of the angular frequency, is given by
\begin{equation}
  \eqlabel{Schuster}
  C(\omega) = \frac{1}{N}\left| \sum_{k=1}^N y_k e^{i\omega t_k} \right|^2,
\end{equation}
and can be computed very efficiently using the Fast Fourier Transform.

@@

While the important features of least squares frequency estimation via Lomb-Scargle periodograms has been discussed elsewhere, we will present a brief introduction to the subject in the following section.
The matrix-based formalism we develop here will make clear how the method can be extended to more sophisticated models, including the multiband method.

@@

In this section we present a quick quantitative introduction to the least squares fitting formulation of the normalized periodogram of \eq{LombScargle}. We denote $N$ observed data points as
\begin{equation}
  D = \{t_k, y_k, \sigma_k\}_{k=1}^N
\end{equation}
where $t_k$ is the time of observation, $y_k$ is the observed value (typically a magnitude), and $\sigma_k$ describes the Gaussian errors on each value. Without loss of generality we will assume that the data $y_k$ are centered such that the measurements within each band satisfy
\begin{equation}
  \eqlabel{ycentered}
  \frac{\sum_k w_ky_k}{\sum_k w_k} = 0
\end{equation}
where the weights are $w_k = \sigma_k^{-2}$.

@@

The matrix-based formulation of \eq{LombScargle2}, though slower than the Fourier-derived formulation, is a more general expression and allows several advantages:
\begin{enumerate}
  \item It is trivially extended to heteroscedastic and/or correlated measurement noise in the data $y_k$ through appropriate modification of the {\it noise covariance matrix} $\Sigma$
  \item It is trivially extended to more sophisticated linear models by appropriately modifying the {\it design matrix} $X_\omega$.
  \item It is trivially extended to include Tikhonov/L2-regularization terms (see \S \ref{regularization} for more details)  by adding an appropriate diagonal term to the {\it normal matrix} $X_\omega^T\Sigma^{-1}X_\omega$.
\end{enumerate}
In the remainder of this section, we will explore a few of these modifications and how they affect the periodogram and resulting model fits.

@@

\subsection{Regularized Models \label{regularization}}

@@

The previous sections raise the question: how complicated a model should we use? We have seen that as we add more terms to the fit, the model will come closer and closer to the observed data.  In the extreme case, when the number of model parameters equals the number of data points, the model can fit the data {\it exactly} regardless of frequency and the periodogram will be everywhere unity (though in most cases, numerical inaccuracies in the fit prevent a truly perfect fit). For very high-order models, the fit becomes very sensitive to the noise in the data, and we say we have {\it over-fit} the data. This can be addressed by explicitly truncating the series, but we can also use a {\it regularization} term to mathematically enforce a less complicated model.

@@

\section{A Multiple-Band Model}
\sectlabel{multiband}
In this section we will apply what was learned in the previous sections to construct the {\it multiband periodogram} which can flexibly account for heterogeneous sources of data for a single object.
To compute such a periodogram, we will take advantage of the easy extensibility of the matrix formalism which led to our generalizations above.
The model contains the following features:
\begin{enumerate}
  \item An $N_{base}$-term truncated Fourier fit which models a latent parameter, which we'll call the ``overall variability''.
  \item A set of $N_{band}$-term truncated Fourier fits, each of which models the residual of a single band from this overall variability.
\end{enumerate}
The total number of parameters for $K$ filters is then $M_K = (2N_{base} + 1) + K(2N_{band} + 1)$. As a result, for each band $k$ we have the following model of the observed magnitudes:
\begin{eqnarray}
  \eqlabel{multiband_model}
  y_k(t|\omega,\theta) = &\theta_0 + \sum_{n=1}^{M_{base}} \left[\theta_{2n - 1}\sin(n\omega t) + \theta_{2n}\cos(n\omega t)\right]& +\nonumber\\ 
  &\theta^{(k)}_0 + \sum_{n=1}^{M_{band}} \left[\theta^{(k)}_{2n - 1}\sin(n\omega t) + \theta^{(k)}_{2n}\cos(n\omega t)\right].&
\end{eqnarray}
The important feature of this model is that {\it all bands} share the same base parameters $\theta$, while their offsets $\theta^{(k)}$ are determined individually.

We can construct the normalized periodogram for this model by building a sparse design matrix with $M_K$ columns. Each row corresponds to a single observation through a single band. Columns corresponding to the base model and the matching observation band will have nonzero entries; all other columns will be filled with zeros. For example, the $N_{base}=1$ and $N_{band}=0$ model corresponds to one with a simple single-term periodic base frequency, and an independent constant offset term in each band. The associated design matrix depends on the particular data, but will look similar to this:
\begin{equation}
X_\omega^{(1,0)} = \left[
\begin{array}{cccccccc}
1 & \sin(\omega t_1) & \cos(\omega t_1) & 1 & 0 & 0 & 0 & 0\\
1 & \sin(\omega t_2) & \cos(\omega t_2) & 0 & 0 & 0 & 0 & 1\\
1 & \sin(\omega t_3) & \cos(\omega t_3) & 0 & 0 & 0 & 1 & 0\\
\vdots & \vdots & \vdots & & & \vdots & &\\
1 & \sin(\omega t_N) & \cos(\omega t_N) & 0 & 0 & 1 & 0 & 0\\
\end{array}
\right]
\end{equation}
Here the nonzero entries of the final five columns are binary flags indicating the $(u, g, r, i, z)$-band of the given observation: for this example, the first row is a $u$-band measurement, the second is a $z$-band, the third is a $i$-band, etc.

@@

\subsection{Relationship of Multiband and Single-band approaches}
\sectlabel{relationship}
The {\it multi-phase} $(N_{base}=0, N_{band}=1)$ model turns out to be a particularly special case.
The base model is a simple global offset which is degenerate with the offsets in each band, so that the design matrix $X_\omega$ can be straightforwardly rearranged as block-diagonal. A block-diagonal design matrix in a linear model indicates that components of the model are being solved independently: here these independent components amount to the single-band floating-mean model from \sect{floating_mean}, fit independently for each of the $K$ bands. This particular multiband model can give us insight into the relationship between single-band and multiband approaches.

@@

The lower panels of \fig{multiband_sim} show the same 60 nights of data, except with only a {\it single} band observation recorded each night. The lower-left panel shows the observations as a function of phase, and the lower-right panels show the periodograms derived from the data. With only 12 observations for each indiviudal band, it is clear that there is not enough data to accurately determine the period within each single band. The shared-phase multiband approach (i.e. $(N_{base}=1,N_{band}=0)$), shown in the lower-right panel, makes use of all the information in a single model, and clearly recovers the true frequency of 0.622 days.

This shared-phase $(1,0)$ model is only one of the possible multiband options, however: \fig{multiband_models} shows multiband fits to this data for models with various choices of $(N_{base},N_{band})$. We see here many of the characteristics noted above for single-band models: as discussed in \sect{multiterm}, increasing the number of Fourier terms leads to power at multiples of the fundamental period. Additionally, increasing the model complexity (roughly indexed by the effective number of free parameters $M^{eff}$) tends to increase the background level of the periodogram, potentially obscuring significant peaks. For this reason, models with $N_{base} > N_{band}$ are the most promising: they allow a flexible fit with minimal model complexity. Motivated by this, in the next section we'll apply the simplest of this class of models, the $(1, 0)$ shared-phase model, to data from the Stripe 82 of the Sloan Digital Sky Survey.

\section{Application to Stripe 82 RR Lyrae}
\sectlabel{stripe82}
Stripe 82 is a three hundred square degree equatorial region of the sky which was repeatedly imaged through multiple band-passes during phase II of the Sloan Digital Sky Survey \citep[SDSS II, see][]{Sesar2007}.
Here we consider the SDSS II observations of 483 RR Lyrae stars compiled and studied by S10.
S10 determined periods for these stars based on emperically-derived light curve templates. Because the template-fitting method is extremely computationally intensive, S10 first determined candidate periods by taking the top 5 results of the Supersmoother \citep{Reimann94} algorithm applied to the $g$-band; template fits were then performed at each candidate period and the period with the best template fit was reported as the true period. In this section, we make use of this dataset to quantitatively evaluate the effectiveness of the multiband periodogram approach.

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{fig07a.pdf}
  \includegraphics[width=\textwidth]{fig07b.pdf}
  \caption{
    Comparison of the Multiband algorithm and single-band supersmoother algorithm on 483 well-sampled RR Lyrae light curves from Stripe 82.
    The upper panels show a representative lightcurve and periodogram fits, while the bottom panels compare the derived periods to the template-based periods reported in S10.
    Shown for reference are the beat aliases (dotted lines) and the multiplicative alias (dashed lines): numbers along the top and right edges of the panels indicate the number of points aligned with each trend.
    The single-band supersmoother algorithm tends to err toward multiplicative aliases, while the multiband algorithm tends to err toward beat frequency aliases.
    Both methods find the correct period among the top 5 significant peaks around 99\% of the time.
  } 
  \figlabel{compare_periods}
\end{figure}

The full S10 RR Lyrae dataset consists of 483 objects with an average of 55 observations in each of the five SDSS $ugriz$ bands spread over just under ten years.  In the upper panels of \fig{compare_periods} we show the observed data for one of these objects, along with the periodogram derived with the single-band supersmoother algorithm and the shared-phase $(0, 1)$-multiband algorithm\footnote{The supersmoother ``periodogram'' $P_{SS}$ is constructed from the minimum sum of weighted model residuals $\bar{r}_{min}$ in analogy with \eq{chi2PN}: $P_{SS}(\omega) = 1 - \bar{r}_{min}(\omega) / \bar{r}_0$, where $\bar{r}_0$ is the residual around a constant model.}. Here we have a case which is analogous to that shown for simulated data in the top panels of \fig{multiband_sim}: each band has enough data to easily locate candidate peaks, the best of which is selected via the S10 template-fitting procedure.

@@

The best-fit supersmoother period matches the S10 period in 89\% of cases (421/483), while the best-fit multiband period matches the S10 period in 79\% of cases (382/483). The modes of failure are instructive: when the supersmoother model misses the S10 period, it tends to land on a multiplicative alias (i.e. the dashed line). This is due to the flexibility of the supersmoother model: a doubled period spreads the points out, leading to fewer constraints in each neighborhood and thus a smaller average residual around model. In other words, the SuperSmoother tends to over-fit data which is sparsely-sampled. On the other hand, when the multiband model misses the S10 period, it tends to land on a beat alias between the S10 period and the 1-day observing cadence (i.e. the dotted lines). This is due to the fact that the single-frequency periodic model is biased, and significantly under-fits the data: it cannot distinguish residuals due to underfitting from residuals due to window function effects.

@@

Above we saw that the multiband model is comparable to methods from the literature for densely-sampled data. Where we expect the multiband approach to gain an advantage is when the data are sparsely sampled, with observations through only a single band at each observation. To simulate this, we reduce the size of the Stripe 82 RR Lyrae dataset by a factor of 5, keeping only a single band of imaging each night: an average of 11 observations of each object per band. This is much closer to the type of data which will be available in future multiband time-domain surveys.

@@

The lower panels of \fig{compare_periods_reduced} show the relationship between the S10 periods (based on the full dataset) and the periods derived with each model from this reduced dataset. It is clear that the supersmoother model is simply over-fitting noise with this few data points: the top period matches S10 in only 32\% of cases (compared to 87\% with the full dataset), and the top 5 periods contain the S10 period only 45\% of the time. The failure mode is much less predictable as well: rather than being clustered near aliases, most of the period determinations are scattered seemingly randomly around the space.

@@

Another potentially fruitful way forward in the multiband case would be to make use of physically-motivated priors in the light-curve shape.
This approach could allow the model bias to be decreased without a commensurate increase in model complexity, which is what causes poor performance in the case of sparsely-sampled noisy data.
As an example of such a physically-motivated prior, consider that the paths of RR Lyrae stars through color-color and color-magnitude space are constrained by known astrophysical processes in the structure of the stars \citep[e.g., see Fig. 5 in][]{Szabo2014}. Making use of this information could help break degeneracies in period determination with higher-order models.

@@

One potential way forward is to move away from nonparametric models like supersmoother and lomb-scargle to template-fitting methods such as those used in S10.
Indeed, such methods perform well even for sparsely-sampled multiband data such as those from the PanSTARRS survey (B. Sesar, private communication); the primary drawback is that such blind template fits are computationally extremely expensive: they involve nonlinear optimizations over each of several hundred candidate templates at each of tens of thousands of candidate frequencies.
Thus the template-fitting method, though extremely accurate, in practice requires several hours of computing time for a well-sampled period grid for a single source, compared to several seconds for the multiband periodogram proposed here.
Note that several hours per object is orders-of-magnitude too slow in the case of LSST; to estimate periods for a billion stars on a 1000-core machine in a year requires a compute-time budget of only 30 seconds per light curve.

Because of the computational expense of the pure template-fitting method, when working with SDSS II data S10 performed a first-pass with a single-band supersmoother to establish candidate periods, which were in turn evaluated with template-fitting approach.
Here we show that such a hybrid strategy combining the multiband periodogram and the S10 template fits will be useful for determining periodicity of variables in early LSST data releases, greatly improving on the outlook presented in \citet{Oluseyi12}.
Rather than doing the full analysis including the final template fits, we will focus on the ability of the multiband periodogram to quickly provide suitable candidate periods, and here assume that the S10 template algorithm will then select or reject the optimial period from this set.

We suggest the following procedure for determining periods in future multiband datasets:
\begin{enumerate}
   \item As a first pass, find a set of candidate frequencies using the multiband periodogram. This is a fast linear optimization that can be trivially parallelized.
   \item Within these candidate frequencies, use the much more costly template-fitting procedure to choose the optimal period from among the handful of candidates.
   \item Compute a goodness-of-fit statistic for the best-fit template to determine whether the fit is suitable; if not, then apply the template-fitting procedure across the full period range.
\end{enumerate}
Here we briefly explore simulated LSST observations of RR Lyrae stars in order to gauge the effectiveness of the first step in this strategy; the effectiveness of the template-fitting step will be explored further in future work.

@@

\fig{LSST_sims} shows the fraction of stars where this period matching criterion is met as a function of $g$-band magnitude and subset of LSST data.
The solid lines show the multiband results; the dashed lines show the single-band supersmoother results; and the shading helps guide the eye for the sake of comparison.
Because of our relaxed matching criteria, the dashed line here is everywhere higher than the \citet{Oluseyi12} results (compare to Figure 15 in that work); thus the supersmoother result here can be considered a best-case scenario for single-band fits.
Nevertheless, without fail the multiband result is superior to the single-band result. The improvement is most apparent for faint stars, where the greater model flexibility of the supersmoother causes it to over-fit the noisy data.

These results point to much more promising prospects for science with variable stars than previously reported.
In particular, even with only six months of LSST data, we can expect to correctly identify the periods for over 60\% of stars brighter than $g=22$; with the first two years of LSST observations, this increases to nearly 100\%.
Part of this improvement is due to the performance of the shared-phase multiband model with noisy data, and part of this improvement is due to the relaxed period-matching constraints enabled by the hybrid approach of non-parametric and template-based period determination.

>>>>

\begin{abstract}
This paper introduces the {\it multiband periodogram}, a general extension of the well-known 
Lomb-Scargle approach for detecting periodic signals in time-domain data. In addition to 
advantanges of the Lomb-Scargle method such as treatment of non-uniform sampling and
heteroscedastic errors, the multiband periodogram significantly improves period finding for randomly 
sampled multiband light curves (e.g., Pan-STARRS, DES and LSST). The light curves in 
each band are modeled as arbitrary truncated Fourier series, with the period and phase 
shared across all bands. The key aspect is the use of Tikhonov regularization which drives
most of the variability into the so-called base model common to all bands, while 
fits for individual bands describe residuals relative to the base model and typically require
lower-order Fourier series. This decrease in the required number of fit parameters is the
main reason for improved performance. We use simulated light curves 
and randomly subsampled SDSS Stripe 82 data to demonstrate the superiority of this 
method compared to other methods from the literature, and find that this method will be able to
efficiently determine the correct period in the majority of LSST's bright RR Lyrae stars with
as little as six months of LSST data.
A Python implementation of this method, along with code to fully reproduce the results
reported here, is available on GitHub.
\end{abstract}

@@

The proposed method is essentially a generalization of the Lomb-Scargle method to 
multiband case. The light curves in  each band are modeled as arbitrary truncated Fourier series, 
with the period, and optionally the phase, shared across all bands. The key aspect enabling this approach is the use of Tikhonov regularization 
(discussed in detail in \sect{regularization}) which drives most of the variability into the so-called {\it base 
model} common to all bands, while fits for individual bands describe residuals relative to the base model 
and typically require lower-order Fourier series. This decrease in the required number of fit parameters is the
main reason for improved performance.

The remainder of the paper is organized as follows:
in \sect{brief_overview} we provide a brief review of least-squares periodic fitting, and in \sect{matrix_formalism} derive the matrix-based formalism for single-band periodic analysis used through the rest of this work.
\sect{extending_periodogram} introduces several extensions and generalizations to the single-band model that the matrix formalism makes possible, including floating mean models, truncated Fourier models, and regularized models.
In \sect{multiband}, we use the ideas behind these extensions to motivate the {\it multiband periodogram}, and show some examples of its use on simulated data.
In \sect{stripe82} we apply this method to measurements of 483 RR Lyrae stars first explored by \citet[][hereafter S10]{Sesar2010}, and in \sect{LSST} explore the performance of the method for simulated observations from the LSST survey.
We conclude in \sect{discussion}.

\section{Brief Overview of Periodic Analysis}
\sectlabel{brief_overview}

The detection and quantification of periodicity in time-varying signals is an important area of data analysis within modern time-domain astronomical surveys.
For evenly-spaced data, the {\it periodogram}, a term coined by \citet{Schuster98}, gives a quantitative measure of the periodicity of data as a function of the angular frequency $\omega$. For data $\{y_k\}_{k=1}^N$ measured at equal intervals $t_k = t_0 + k\Delta t$, Schuster's periodogram, which measures the spectral power as a function of the angular frequency, is given by
\begin{equation}
  \eqlabel{Schuster}
  C(\omega) = \frac{1}{N}\left| \sum_{k=1}^N y_k e^{i\omega t_k} \right|^2,
\end{equation}
and can be computed very efficiently using the Fast Fourier Transform.

@@

While the important features of least squares frequency estimation via Lomb-Scargle periodograms have been discussed elsewhere, we will present a brief introduction to the subject in the following section.
The matrix-based formalism we develop here will make clear how the method can be extended to more sophisticated models, including the multiband periodogram proposed in this work.

@@

In this section we present a brief quantitative introduction to the least squares fitting formulation of the normalized periodogram of \eq{LombScargle}. We denote $N$ observed data points as
\begin{equation}
  D = \{t_k, y_k, \sigma_k\}_{k=1}^N
\end{equation}
where $t_k$ is the time of observation, $y_k$ is the observed value (typically a magnitude), and $\sigma_k$ describes the Gaussian errors on each value. Without loss of generality we will assume that the data $y_k$ are centered such that the measurements within each band satisfy
\begin{equation}
  \eqlabel{ycentered}
  \frac{\sum_k w_ky_k}{\sum_k w_k} = 0
\end{equation}
where the weights are $w_k = \sigma_k^{-2}$.

@@

The matrix-based formulation of \eq{LombScargle2}, though slower than the Fourier-derived formulation, is a more general expression and allows several advantages:
\begin{enumerate}
  \item It is trivially extended to heteroscedastic and/or correlated measurement noise in the data $y_k$ through appropriate modification of the {\it noise covariance matrix} $\Sigma$.
  \item It is trivially extended to more sophisticated linear models by appropriately modifying the {\it design matrix} $X_\omega$.
  \item It is trivially extended to include Tikhonov/L2-regularization terms (see \sect{regularization} for more details)  by adding an appropriate diagonal term to the {\it normal matrix} $X_\omega^T\Sigma^{-1}X_\omega$.
\end{enumerate}
In the remainder of this section, we will explore a few of these modifications and how they affect the periodogram and resulting model fits.

@@

\subsection{Regularized Models}
\sectlabel{regularization}

@@

The previous sections raise the question: how complicated a model should we use? We have seen that as we add more terms to the fit, the model will come closer and closer to the observed data.  In the extreme case, when the number of model parameters equals the number of data points, the model can fit the data {\it exactly} regardless of frequency and the periodogram will be everywhere unity (though in most cases, numerical inaccuracies prevent a truly perfect fit). For very high-order models, the fit becomes very sensitive to the noise in the data, and we say we have {\it over-fit} the data. This can be addressed by explicitly truncating the series, but we can also use a {\it regularization} term to mathematically enforce a less complicated model.

@@

\section{A Multiple-Band Model}
\sectlabel{multiband}
In this section we will apply what was learned in the previous sections to construct the {\it multiband periodogram} which can flexibly account for heterogeneous sources of data for a single object.
To compute such a periodogram, we will take advantage of the easy extensibility of the matrix formalism which led to our generalizations above.
The multiband model contains the following features:
\begin{enumerate}
  \item An $N_{base}$-term truncated Fourier fit which models a latent parameter, which we'll call the ``overall variability''.
  \item A set of $N_{band}$-term truncated Fourier fits, each of which models the residual of a single band from this overall variability.
\end{enumerate}
The total number of parameters for $K$ filters is then $M_K = (2N_{base} + 1) + K(2N_{band} + 1)$. As a result, for each band $k$ we have the following model of the observed magnitudes:
\begin{eqnarray}
  \eqlabel{multiband_model}
  y_k(t|\omega,\theta) = &\theta_0 + \sum_{n=1}^{M_{base}} \left[\theta_{2n - 1}\sin(n\omega t) + \theta_{2n}\cos(n\omega t)\right]& +\nonumber\\ 
  &\theta^{(k)}_0 + \sum_{n=1}^{M_{band}} \left[\theta^{(k)}_{2n - 1}\sin(n\omega t) + \theta^{(k)}_{2n}\cos(n\omega t)\right].&
\end{eqnarray}
The important feature of this model is that {\it all bands} share the same base parameters $\theta$, while their offsets $\theta^{(k)}$ are determined individually.

We can construct the normalized periodogram for this model by building a sparse design matrix with $M_K$ columns. Each row corresponds to a single observation through a single band. Columns corresponding to the base model and the matching observation band will have nonzero entries; all other columns will be filled with zeros. For example, the $N_{base}=1$ and $N_{band}=0$ model corresponds to one with a simple single-term periodic base frequency, and an independent constant offset term in each band. The associated design matrix depends on the particular data, but will look similar to this:
\begin{equation}
X_\omega^{(1,0)} = \left[
\begin{array}{cccccccc}
1 & \sin(\omega t_1) & \cos(\omega t_1) & 1 & 0 & 0 & 0 & 0\\
1 & \sin(\omega t_2) & \cos(\omega t_2) & 0 & 0 & 0 & 0 & 1\\
1 & \sin(\omega t_3) & \cos(\omega t_3) & 0 & 0 & 0 & 1 & 0\\
\vdots & \vdots & \vdots & & & \vdots & &\\
1 & \sin(\omega t_N) & \cos(\omega t_N) & 0 & 0 & 1 & 0 & 0\\
\end{array}
\right]
\end{equation}
Here the nonzero entries of the final five columns are binary flags indicating the $(u, g, r, i, z)$-band of the given observation: for this example, the first row is a $u$-band measurement, the second is a $z$-band, the third is a $i$-band, etc., as indicated by the position of the nonzero matrix element within the row.

@@

\subsection{Relationship of Multiband and Single-band approaches}
\sectlabel{relationship}
The {\it multi-phase} $(N_{base}=0, N_{band}=1)$ model turns out to be a particularly special case.
Here the base model is a simple global offset which is degenerate with the offsets in each band, so that the design matrix $X_\omega$ can be straightforwardly rearranged as block-diagonal.
A block-diagonal design matrix in a linear model indicates that components of the model are being solved independently: here these independent components amount to the single-band floating-mean model from \sect{floating_mean}, fit independently for each of the $K$ bands.
This particular multiband model can give us insight into the relationship between single-band and multiband approaches.

@@

The lower panels of \fig{multiband_sim} show the same 60 nights of data, except with only a {\it single} band observation recorded each night.
The lower-left panel shows the observations as a function of phase, and the lower-right panels show the periodograms derived from the data.
With only 12 observations for each indiviudal band, it is clear that there is not enough data to accurately determine the period within each single band. The shared-phase multiband approach (i.e. $N_{base}=1,N_{band}=0$), shown in the lower-right panel, fits a single model to the full data and clearly recovers the true frequency of 0.622 days.

This shared-phase $(1,0)$ model is only one of the possible multiband options, however: \fig{multiband_models} shows multiband fits to this data for models with various choices of $(N_{base},N_{band})$.
We see here many of the characteristics noted above for single-band models: as discussed in \sect{multiterm}, increasing the number of Fourier terms leads to power at multiples of the fundamental period, and increased model complexity (roughly indexed by the effective number of free parameters $M^{eff}$) tends to increase the background level of the periodogram, obscuring significant peaks.
For this reason, models with $N_{base} > N_{band}$ are the most promising: they allow a flexible fit with minimal model complexity. Motivated by this, in the next section we'll apply the simplest of this class of models, the $(1, 0)$ shared-phase model, to data from the Stripe 82 of the Sloan Digital Sky Survey.

\section{Application to Stripe 82 RR Lyrae}
\sectlabel{stripe82}
Stripe 82 is a three hundred square degree equatorial region of the sky which was repeatedly imaged through multiple band-passes during phase II of the Sloan Digital Sky Survey \citep[SDSS II, see][]{Sesar2007}.
Here we consider the SDSS II observations of 483 RR Lyrae stars compiled and studied by S10, in which periods for these stars were determined based on emperically-derived light curve templates.
Because the template-fitting method is extremely computationally intensive, S10 first determined candidate periods by taking the top 5 results of the Supersmoother \citep{Reimann94} algorithm applied to the $g$-band; template fits were then performed at each candidate period and the period with the best template fit was reported as the true period. In this section, we make use of this dataset to quantitatively evaluate the effectiveness of the multiband periodogram approach.

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{fig07a.pdf}
  \includegraphics[width=\textwidth]{fig07b.pdf}
  \caption{
    Comparison of the Multiband algorithm and single-band supersmoother algorithm on 483 well-sampled RR Lyrae light curves from Stripe 82.
    The upper panels show a representative lightcurve and periodogram fits, while the bottom panels compare the derived periods to the template-based periods reported in S10.
    Shown for reference are the beat aliases (dotted lines) and the multiplicative alias (dashed lines): numbers along the top and right edges of the panels indicate the number of points aligned with each trend.
    The single-band supersmoother model tends to err toward multiplicative aliases, while the multiband model tends to err toward beat frequency aliases.
    Both methods find the correct period among the top 5 significant peaks around 99\% of the time.
  } 
  \figlabel{compare_periods}
\end{figure}

The full S10 RR Lyrae dataset consists of 483 objects with an average of 55 observations in each of the five SDSS $ugriz$ bands spread over just under ten years.  In the upper panels of \fig{compare_periods} we show the observed data for one of these objects, along with the periodogram derived with the single-band supersmoother model and the shared-phase $(0, 1)$-multiband model\footnote{The supersmoother ``periodogram'' $P_{SS}$ is constructed from the minimum sum of weighted model residuals $\bar{r}_{min}$ in analogy with \eq{chi2PN}: $P_{SS}(\omega) = 1 - \bar{r}_{min}(\omega) / \bar{r}_0$, where $\bar{r}_0$ is the mean absolute residual around a constant model.}. Here we have a case which is analogous to that shown for simulated data in the top panels of \fig{multiband_sim}: each band has enough data to easily locate candidate peaks, the best of which is selected via the S10 template-fitting procedure.

@@

The best-fit supersmoother period matches the S10 period in 89\% of cases (421/483), while the best-fit multiband period matches the S10 period in 79\% of cases (382/483). The modes of failure are instructive: when the supersmoother model misses the S10 period, it tends to land on a multiplicative alias (i.e. the dashed line). This is due to the flexibility of supersmoother: a doubled period spreads the points out, leading to fewer constraints in each neighborhood and thus a smaller average residual around model. In other words, the SuperSmoother tends to over-fit data which is sparsely-sampled. On the other hand, when the multiband model misses the S10 period, it tends to land on a beat alias between the S10 period and the 1-day observing cadence (i.e. the dotted lines). This is due to the fact that the single-frequency periodic model is biased, and significantly under-fits the data: it cannot distinguish residuals due to underfitting from residuals due to window function effects.

@@

Above we saw that the multiband model is comparable to methods from the literature for densely-sampled data. Where we expect the multiband approach to gain an advantage is when the data are sparsely sampled, with data through only a single band at each observation time. To simulate this, we reduce the size of the Stripe 82 RR Lyrae dataset by a factor of 5, keeping only a single band of imaging each night: an average of 11 observations of each object per band. This is much closer to the type of data which will be available in future multiband time-domain surveys.

@@

The lower panels of \fig{compare_periods_reduced} show the relationship between the S10 periods (based on the full dataset) and the periods derived with each model from this reduced dataset. It is clear that the supersmoother model is simply over-fitting noise with this few data points: the top period matches S10 in only 32\% of cases (compared to 87\% with the full dataset), and the top 5 periods contain the S10 period only 45\% of the time. The failure mode is much less predictable as well: rather than being clustered near aliases, most of the period determinations are scattered seemingly randomly around the parameter space.

@@

One potentially fruitful avenue of research which we do not study here is the application of L1 regularization (also known as Lasso regression) to the higher-order periodogram.
This is similar in spirit to the Tikhonov regularization discussed in \sect{regularization}, but tends toward sparsity in the model parameters \citep[see, e.g.][for a discussion]{ICVG2014}.
Such an approach could provide a useful tradeoff between model complexity and bias in the case of higher-order truncated Fourier models.

Another potentially fruitful way forward in the multiband case would be to define and make use of physically-motivated priors in the light-curve shape.
This approach could allow the model bias to be decreased without a commensurate increase in model complexity, which is what causes poor performance in the case of sparsely-sampled noisy data.
As an example of such a physically-motivated prior, consider that the paths of RR Lyrae stars through color-color and color-magnitude space are constrained by known astrophysical processes in the structure of the stars \citep[e.g., see Fig. 5 in][]{Szabo2014}. Making use of this information could help break degeneracies in period determination with higher-order models.

@@

One potential remedy is to move away from general models like supersmoother and lomb-scargle to specific template-fitting methods such as those used in S10.
Indeed, such methods perform well even for sparsely-sampled multiband data such as those from the PanSTARRS survey; the primary drawback is that such blind template fits are computationally extremely expensive: they involve nonlinear optimizations over each of several hundred candidate templates at each of tens of thousands of candidate frequencies (B. Sesar, private communication).
Thus the template-fitting method, though extremely accurate, in practice requires several hours of CPU time for a well-sampled period grid for a single source (compared to several seconds for the multiband periodogram proposed here).
Note that several hours per object is orders-of-magnitude too slow in the case of LSST; to estimate periods for a billion stars on a 1000-core machine in a year requires a compute-time budget of only 30 seconds per light curve.

Because of the computational expense of the pure template-fitting method, when working with SDSS II data S10 performed a first-pass with a single-band supersmoother to establish candidate periods, which were in turn evaluated with template-fitting approach.
Here we show that such a hybrid strategy combining the multiband periodogram and the S10 template fits will be useful for determining periodicity of variables in early LSST data releases, greatly improving on the outlook presented in \citet{Oluseyi12}.

We suggest the following procedure for determining periods in future multiband datasets:
\begin{enumerate}
   \item As a first pass, find a set of candidate frequencies using the multiband periodogram. This is a fast linear optimization that can be trivially parallelized.
   \item Within these candidate frequencies, use the much more costly template-fitting procedure to choose the optimal period from among the handful of candidates.
   \item Compute a goodness-of-fit statistic for the best-fit template to determine whether the fit is suitable; if not, then apply the template-fitting procedure across the full period range.
\end{enumerate}
Here we briefly explore simulated LSST observations of RR Lyrae stars in order to gauge the effectiveness of the first step in this strategy; the effectiveness of the template-fitting step will be explored further in future work.
Rather than doing the full analysis including the final template fits, we will focus on the ability of the multiband periodogram to quickly provide suitable candidate periods under the assumption that the S10 template algorithm will then select or reject the optimial period from this set.

\subsection{LSST Simulations}

@@

\fig{LSST_sims} shows the fraction of stars where this period matching criterion is met as a function of $g$-band magnitude and subset of LSST data.
The solid lines show the multiband results; the dashed lines show the single-band supersmoother results; and the shading helps guide the eye for the sake of comparison.
Because of our relaxed matching criteria, even the single-band supersmoother results here are much more optimistic than the \citet{Oluseyi12} results (compare to Figure 15 in that work): the supersmoother result here can be considered a best-case scenario for \adhoc{} single-band fits.
Without fail, the multiband result exceeds this best-case single-band result; the improvement is most apparent for faint stars, where the greater model flexibility of the supersmoother causes it to over-fit the noisy data.

These results point to much more promising prospects for science with variable stars than previously reported.
In particular, even with only six months of LSST data, we can expect to correctly identify the periods for over 60\% of stars brighter than $g=22$; with the first two years of LSST observations, this increases to nearly 100\%.
Part of this improvement is due to the performance of the shared-phase multiband model with noisy data, and part of this improvement is due to the relaxed period-matching constraints enabled by the hybrid approach of periodogram-based and template-based period determination.
